Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-18 second-draft/pnas-word-order-si.tex
SI

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parsability.
The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
%We previously carried out language optimization for $\lambda = 1$, and obtained essentially identical predictions for word order correlations, see Table REF.
We note that each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions (Table~\ref{tab:all-predictions-1}).
%Indeed, the same holds for all of the 38 syntactic relations in UD  (Table~\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})
%We additionally report results with $\lambda = 0$, i.e., placing all weight on parseability.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}).
We also verified the robustness of our results to small variations of $\lambda$, by reporting results with $\lambda=1$ in Table~\ref{table:corr-resu}.

@@

Within each language, we estimate the Pareto curve based on the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section~\ref{sec:lambda}), (2) parseability, and (3) predictability.
%We then estimated the overall average Pareto curve, as shown in Figure (REF), by averaging the z-transformed points (1), (2), (3).

@@

\begin{figure} 
	\begin{center}
	\includegraphics[width=0.7\textwidth]{../results/correlations/figures/pred-eff-pred-pars-families-2.pdf} % figures/pred-eff-families-2.pdf
	\end{center}
	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by optimizing for efficiency, predictability, parseability (right). Dark blue: Verb patterner \emph{precedes} object patterner (English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...). White cells indicate that the relation is not annotated in the dataset for the given language.}\label{fig:per-lang}
%	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by efficiency optimization (right). Dark blue: Verb patterner \emph{precedes} object patterner(English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...) \textcolor{red}{add other functions to this plot}.}\label{fig:per-lang}
\end{figure}

@@

Further, we report results for grammars optimized for dependency length minimization (DLM, see Section~\ref{sec:DLM} for discussion).
This has been suggested as a property of efficient word order in prior work, and, indeed, we found that DLM is predicted by efficiency optimization (Section REF).
Results in Table~\ref{tab:all-predictions-1} also show that optimizing for DLM makes predictions similar to efficiency optimization. See Section REF.

@@

    \multicolumn{7}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}

@@

@@

\subsection{Other Experiments}
In Table~\ref{table:corr-resu} we report the results of the simulations we did before the main experiment reported in the main paper (left and center), together with their preregistrations, besides the results from the main experiment (right).
These experiments all had the same setup described in Section (REF), which was fixed before starting simulations; differences are that (1) the first simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.
\textcolor{red}{TODO also the aggregate posterior}

@@

& $\lambda=1.0$\footnote{\url{https://aspredicted.org/blind.php?x=bg35x7}. Note that we report the locality results as `DLM' in Table REF.} 

@@

@@

These quatities are estimated using two neural models, as described in Section X:

@@

%We initialize all values $a_\tau := 0.5$, $b_\tau := 0$ (except for the \emph{obj} dependency, for which we fix $a_\tau$ to 0 or 1, see Section (REF)).

@@

\paragraph{Improved Unbiased Gradient Estimator}
We employ two common variance reduction methods to improve the estimator~(\ref{eq:dtheta}), while keeping it unbiased.
%For dependency length and 
% (and its head, in the case of dependency length)
For predictability, note that the loss incurred for a specific word only depends on ordering decisions made up to that word.
We represent the process of linearizing a tree as a dynamic stochastic computation graph, and use these independence properties to apply the method described in \citet{schulman2015gradient} to obtain a version of~(\ref{eq:dtheta}) with lower variance.
Second, we use a word-dependent moving average of recent per-word losses as control variate \cite{williams1992simple} (the word's surprisal in the case of predictability, and the negative log-probability of the correct head and relation label in the case of parseability).
These two methods reduce the variance of the estimator and thereby increase the speed of optimization and reduce training time, without biasing the results.
For numerical stability, we represent $a_\tau \in [0,1]$ via its logit $\in \mathbb{R}$.
Furthermore, to encourage exploration of the parameter space, we add an entropy regularization term \citep{xu2015show} for each Direction Parameter $a_\tau$, which penalizes $a_\tau$ values near $0$ or $1$. The weight of the entropy regularization was chosen together with the other hyperparameters.\footnote{Explored values: 0.0001, 0.001.}
%, adding the negative entropy of the $Bernoulli(a_\tau)$ to the loss $R$. %, weighted with $\alpha$.

@@

We used the CKY algorithm to compute the parsing loss $H[T|L_\theta(T)]$ on the validation partition of the English and Japanese UD corpora, for random and optimized ordering grammars.
Results (Figure~\ref{fig:cky-parser}) show that optimized grammars are more parseable for exact parsing of a simple PCFG.
% TODO get UAS for compatibility with other parts of SI
%We then compared parsability of random and optimized word order grammars on English data.

\begin{figure}
    \centering
    \includegraphics[scale=.7]{../results/cky/cky-parse.pdf} 
	\caption{Parsability loss $H[T|L_\theta(T)]$ (lower is better) computed by a simple CKY parser, for random word order grammars (red) and word order grammars optimized for efficiency (blue).\textcolor{red}{make plot nicer}}
    \label{fig:cky-parser}
\end{figure}

@@

    \includegraphics[scale=.5]{../results/permuted/adversarial-parse-loss-japanese.pdf}
	\caption{Parseability of baseline grammars and grammars optimized for efficiency, in English (top) and Japanese (bottom), measured by parsing loss $H[T|L_\theta(T)]$ (lower is better), for the four distorted orderings, and the actual orderings (`real').}
    \label{fig:distorted-parser}
\end{figure}

@@

        \caption{Parsing loss ($H[T|\mathcal{L}_\theta(T)]$) for optimized (light blue) and random (black) ordering grammar on Czech data, as a function of the fraction of total training data provided.}
    \label{fig:learning-czech}
\end{figure}

\section{Comparing Extracted Grammars to Linguistic Judgments}

@@

We validated our grammars by comparing the dominant orders of six syntactic relations that are also annotated in the World Atlas of Linguistic Structures (WALS, \cite{haspelmath2005world}).
Among the eight Greenbergian correlations that we were able to test, five are annotated in WALS.
In Table~\ref{tab:grammars-wals}, we compare our grammars with WALS on these five relations, and the verb-object relation.

\begin{table}[htb!]
\small{
\begin{center}
\begin{tabular}{l||ll|ll|ll|ll|ll|ll|llllll}
%	   for(dep in c("obj", "lifted_mark", "acl", "nmod", "obl")) {
		   Language 
		   &	\multicolumn{2}{|c|}{Objects} 
		   &	\multicolumn{2}{|c|}{Adpositions} 
		   &	\multicolumn{2}{|c|}{Compl.} 
		   &	\multicolumn{2}{|c|}{Rel.Cl.} 
		   &	\multicolumn{2}{|c|}{Genitive} 
		   &	\multicolumn{2}{|c|}{PP}  \\ \hline\hline
	\input{../results/grammars/comparison-table.tex}
\end{tabular}
\end{center}
}
\caption{Comparing grammars extracted from databases to linguistic judgments in the World Atlas of Linguistic Structures. For each of the six syntactic relation, the first column provides the ordered coded in the extracted grammar; the second column provides the order coded in WALS (DH for dependent-head, HD for head-dependent order). `?' indicates that WALS has no data.
$*$ indicates that WALS does not list a dominant order, which can mean that both orders occur at similar rates in the language, or that insufficient data was available when compiling WALS \citep{dryer2011evidence}.
Finally, `--' indicates that the relation does not occur in the corpus.
}\label{tab:grammars-wals}
\end{table}

@@

We asked whether efficiency optimization predicts dependency length minimization effects.
We first computed dependency length for grammars optimized for efficiency.
We found that 100\% of grammars optimized for efficiency reduce average dependency length compared to baseline grammars ($p < 0.05$, by one-sided $t$-test). \textcolor{red}{TODO}
This suggests that the reduction of dependency length is predicted by efficiency maximization.
Next, we constructed grammars that minimize average dependency length, using the same gradient descent method as we used for efficiency optimization (Section REF).
We expect that such grammars should have shorter dependency length than the real grammars, or grammars optimized for efficiency.
In Figure~\ref{fig:dlm-avg}, we plot the mean dependency length for optimized, real, and baseline orderings.
We find that optimizing grammars for efficiency reduces dependency length to a similar degree as found in the actual orderings in the corpora, almost up to the limit given by directly optimizing for dependency length.
We also plot more detailed results for four languages in Figure~\ref{fig:dlm-4langs}, plotting dependency length as a function of sentence length as reported in prior work \cite{futrell2015largescale}.
Optiziming grammars for efficiency produces dependency lengths similar to those found in the actual orderings.

@@

%Following \cite{futrell2015largescale}, we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
%We optimize grammars using the same stochastic gradient descent method as we use for efficiency optimization (Section REF).
%Indeed, we found that DLM is predicted by efficiency optimization, as discussed in Section REF.
%Results in Table~\ref{tab:all-predictions-1} show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.

>>>>

In our experiments, we chose $\lambda = 0.9$ as a mathematically valid value that puts similar weight on both predictability and parsability.
The computational cost of grammar optimization precluded repeating the experiment for many values of $\lambda$.
We note that each of the eight correlations is predicted by at least parseability or predictability, without any contradictory predictions (Table~\ref{tab:all-predictions-1}).
%Indeed, the same holds for all of the 38 syntactic relations in UD  (Table~\ref{tab:all-predictions-1}-\ref{tab:all-predictions-2})
%We additionally report results with $\lambda = 0$, i.e., placing all weight on parseability.
That is, at $\lambda$ close to its maximal value, the predictions of optimizing the two scoring functions individually add up to the predictions of efficiency optimization.
Small values of $\lambda$ correspond to the case where predictability plays no role, and only parseability is optimized (Table~\ref{tab:all-predictions-1}).
We also verified the robustness of our results to small variations of $\lambda$, by reporting results with $\lambda=1$ in Table~\ref{table:corr-resu}.

@@

Within each language, we estimate the Pareto curve based on the points obtained by optimizing for (1) efficiency (at $\lambda = 0.9$, see Section~\ref{sec:lambda}), (2) parseability, and (3) predictability.

@@

\begin{figure} 
	\begin{center}
	\includegraphics[width=0.7\textwidth]{../results/correlations/figures/pred-eff-pred-pars-families-2.pdf} % figures/pred-eff-families-2.pdf
	\end{center}
	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by optimizing for efficiency, predictability, parseability (right). Dark blue: Verb patterner \emph{precedes} object patterner (English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...). White cells indicate that the relation is not annotated in the dataset for the given language. \textcolor{red}{TODO fix order of 1-8}}\label{fig:per-lang}
%	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by efficiency optimization (right). Dark blue: Verb patterner \emph{precedes} object patterner(English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...) \textcolor{red}{add other functions to this plot}.}\label{fig:per-lang}
\end{figure}

@@

Further, we report results for grammars optimized for dependency length minimization (DLM, see Section~\ref{sec:DLM} for discussion).
This has been suggested as a property of efficient word order in prior work, and, indeed, we found that DLM is predicted by efficiency optimization (Section~\ref{sec:DLM}).
Results in Table~\ref{tab:all-predictions-1} also show that optimizing for DLM makes predictions similar to efficiency optimization. See Section~\ref{sec:DLM}.

@@

@@

 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
\end{tabular}
}

@@

\subsection{Other Experiments}
In Table~\ref{table:corr-resu} we report the results of the simulations we did before the main experiment reported in the main paper (left and center), together with their preregistrations, besides the results from the main experiment (right).
These experiments all had the same setup described in Section~\ref{sec:neural-architectures}, which was fixed before starting simulations; differences are that (1) the first simulation places fully equal weight on parseability and predictability ($\lambda=1.0$), and (2) the final experiment uses three random seeds per grammar.
Results across all three experiments agree; jointly optimizing grammars for parseability and predictability produces all eight correlations.
\textcolor{red}{TODO also the aggregate posterior}

@@

& $\lambda=1.0$\footnote{\url{https://aspredicted.org/blind.php?x=bg35x7}. Note that we report the locality results as `DLM' in Table~\ref{tab:all-predictions-1}, see also Section~\ref{sec:DLM}.} 

@@

We validated the extracted grammars by comparing the dominant orders of six syntactic relations that are also annotated in the World Atlas of Linguistic Structures (WALS, \cite{haspelmath2005world}).
Among the eight Greenbergian correlations that we were able to test, five are annotated in WALS.
In Table~\ref{tab:grammars-wals}, we compare our grammars with WALS on these five relations, and the verb-object relation.

\begin{table}
\small{
\begin{center}
\begin{tabular}{l||ll|ll|ll|ll|ll|ll|llllll}
%	   for(dep in c("obj", "lifted_mark", "acl", "nmod", "obl")) {
		   Language 
		   &	\multicolumn{2}{|c|}{Objects} 
		   &	\multicolumn{2}{|c|}{Adpositions} 
		   &	\multicolumn{2}{|c|}{Compl.} 
		   &	\multicolumn{2}{|c|}{Rel.Cl.} 
		   &	\multicolumn{2}{|c|}{Genitive} 
		   &	\multicolumn{2}{|c|}{PP}  \\ \hline\hline
	\input{../results/grammars/comparison-table.tex}
\end{tabular}
\end{center}
}
\caption{Comparing grammars extracted from databases to linguistic judgments in the World Atlas of Linguistic Structures. For each of the six syntactic relation, the first column provides the ordered coded in the extracted grammar; the second column provides the order coded in WALS (DH for dependent-head, HD for head-dependent order). `?' indicates that WALS has no data.
$*$ indicates that WALS does not list a dominant order, which can mean that both orders occur at similar rates in the language, or that insufficient data was available when compiling WALS \citep{dryer2011evidence}.
Finally, `--' indicates that the relation does not occur in the corpus.
}\label{tab:grammars-wals}
\end{table}

@@

These quatities are estimated using two neural models, as described in Section~\ref{sec:neural-architectures}:

@@

%We initialize all values $a_\tau := 0.5$, $b_\tau := 0$ (except for the \emph{obj} dependency, for which we fix $a_\tau$ to 0 or 1, see Section).

@@

\paragraph{Improved Unbiased Gradient Estimator}
We employ two common variance reduction methods to improve the estimator~(\ref{eq:dtheta}), while keeping it unbiased.
%For dependency length and 
% (and its head, in the case of dependency length)
For predictability, note that the surprisal of a specific word only depends on the preceding words (not on the following words), and thus only depends on ordering decisions made up to that word.
We represent the process of linearizing a tree as a dynamic stochastic computation graph, and use these independence properties to apply the method described in \citet{schulman2015gradient} to obtain a version of~(\ref{eq:dtheta}) with lower variance.
Second, we use a word-dependent moving average of recent per-word losses (the word's surprisal in the case of predictability, and the negative log-probability of the correct head and relation label in the case of parseability) as control variate \cite{williams1992simple}.
These two methods reduce the variance of the estimator and thereby increase the speed of optimization and reduce training time, without biasing the results.
For numerical stability, we represent $a_\tau \in [0,1]$ via its logit $\in \mathbb{R}$.
Furthermore, to encourage exploration of the parameter space, we add an entropy regularization term \citep{xu2015show} for each Direction Parameter $a_\tau$, which penalizes $a_\tau$ values near $0$ or $1$. The weight of the entropy regularization was chosen together with the other hyperparameters.\footnote{Explored values: 0.0001, 0.001.}

@@

We used the CKY algorithm to compute the parsing loss $H[T|\mathcal{L}_\theta(T)]$ on the validation partition of the English and Japanese UD corpora, for random and optimized ordering grammars.
Results (Figure~\ref{fig:cky-parser}) show that optimized grammars are more parseable for exact parsing of a simple PCFG.
% TODO get UAS for compatibility with other parts of SI
%We then compared parsability of random and optimized word order grammars on English data.

\begin{figure}
    \centering
    \includegraphics[scale=.7]{../results/cky/cky-parse.pdf} 
	\caption{Parsability loss $H[T|\mathcal{L}_\theta(T)]$ (lower is better) computed by a simple CKY parser, for random word order grammars (red) and word order grammars optimized for efficiency (blue). We report $H[T|\mathcal{L}_\theta(T)]$ normalized by sentence length. \textcolor{red}{make plot nicer}}
    \label{fig:cky-parser}
\end{figure}

@@

    \includegraphics[scale=.5]{../results/permuted/adversarial-parse-loss-japanese.pdf}
	\caption{Parseability of baseline grammars and grammars optimized for efficiency, in English (top) and Japanese (bottom), measured by parsing loss $H[T|L_\theta(T)]$ (lower is better), for the four distorted orderings, and the actual orderings (`real'). We report $H[T|\mathcal{L}_\theta(T)]$ normalized by sentence length.}
    \label{fig:distorted-parser}
\end{figure}

@@

        \caption{Parsing loss ($H[T|\mathcal{L}_\theta(T)]$, normalized by sentence length) for optimized (light blue) and random (black) ordering grammar on Czech data, as a function of the fraction of total training data provided.}
    \label{fig:learning-czech}
\end{figure}

%\section{Comparing Extracted Grammars to Linguistic Judgments}

@@

@@

We asked whether efficiency optimization predicts dependency length minimization effects.
We first computed dependency length for grammars optimized for efficiency.
We found that 100\% of grammars optimized for efficiency reduce average dependency length compared to baseline grammars ($p < 0.05$, by one-sided $t$-test). \textcolor{red}{TODO}
This suggests that the reduction of dependency length is predicted by efficiency maximization.
Next, we constructed grammars that minimize average dependency length, using the same gradient descent method as we used for efficiency optimization (Section~\ref{sec:optim-eff}).
We expect that such grammars should have shorter dependency length than the real grammars, or grammars optimized for efficiency.
In Figure~\ref{fig:dlm-avg}, we plot the mean dependency length for optimized, real, and baseline orderings.
We find that optimizing grammars for efficiency reduces dependency length to a similar degree as found in the actual orderings in the corpora, almost up to the limit given by directly optimizing for dependency length.
We also plot more detailed results for four languages in Figure~\ref{fig:dlm-4langs}, plotting dependency length as a function of sentence length as reported in prior work \cite{futrell2015largescale}.
Optiziming grammars for efficiency produces dependency lengths similar to those found in the actual orderings.

@@

%Following \cite{futrell2015largescale}, we formalize the dependency length of a sentence as the sum of the distances of all pairs of syntactically related words, where distance is measured by the number of intervening words.
%We optimize grammars using the same stochastic gradient descent method as we use for efficiency optimization (Section ).
%Indeed, we found that DLM is predicted by efficiency optimization, as discussed in Section .
%Results in Table~\ref{tab:all-predictions-1} show that optimizing for DLM makes predictions similar to efficiency optimization, itself a novel result.
